A Responsible Machine Learning Workflow\\\normalsize{With Focus on Interpretable Models, Post-hoc Explanation, and Discrimination Testing}

### 2020-02-04 article-information-2019.tex
fill in accuracy tables
larger, perhaps clearer, XNN charts

% TODO:
% EDITING:
% - Structure ------------------------- x 
% - External links
% - Internal links
%   - Sections/subsections/appendices - x
%   - References ---------------------- x
%   - Tables/figures
%   - footnotes ----------------------- x
% - Abbreviations --------------------- x
% - Bibliography 
%   - Nick/Bryce citation
% - Notation
% - Tables/figures
% - Grammar and spelling
%   - Hyphenation
%   - "metric" vs. "measure"
% CONTENT:
% - sim data: add error metrics, confirm statements about results
% - sim data xnn: update figure: connections, feed-forward direction, make subfigure reference match discrimination measure table
% - mort data: add error metrics, confirm statements about results
% - mort data: update figure: connections, feed-forward direction, make subfigure reference match discrimination measure table
% - make subfigure reference match XNN charts

@@

Fit comparisons between unconstrained and constrained models and XNN interpretabililty results are discussed in Subsections \ref{ssec:c_v_uc_sim} and \ref{ssec:xnn_sim}. As model training and assessment on the simulated data is a rough validation exercise meant to showcase expected results on data with known characteristics, and given that most of the techniques in the proposed workflow are already used widely or have been validated elsewhere, reporting of simulated data results in the main text will focus mostly fit measures for the considered data and model architectures and on the more novel $g^\text{XNN}$ interpretability results. The bulk of the post-hoc explanation and discrimination testing results for the simulated data are left to Appendix \ref{a_sec:res_sim}.

@@

Table \ref{tab:sim_acc} presents a variety of fit metrics for the $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data. $g^\text{GBM}$ exhibits the best performance, but all models exhibit only a small range fit results. Interpretability and explainability benefits of the constrained models appear to come at little cost to overall model performance, or in the case of $g^\text{ANN}$ and $g^\text{XNN}$, no cost at all. $g^\text{XNN}$ actually shows slightly better fit than $g^\text{ANN}$ across all recorded fit metrics. Fit measures that require a probability cutoff are taken at the best F1 threshold for each model.

\begin{table}[H]
	\caption{Fit measures for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data. Arrows indicate the direction of improvement for each measure.}
	\centering
	\begin{tabular}{ccccccccc}
		\toprule
		\textbf{Model} & 
		\href{https://scikit-learn.org/stable/modules/model_evaluation.html#accuracy-score}{\textbf{Accuracy}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve}{\textbf{AUC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/F1_score}{\textbf{F1}}~$\uparrow$ & 
		\href{https://scikit-learn.org/stable/modules/model_evaluation.html#log-loss}{\textbf{Logloss}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Matthews_correlation_coefficient}{\textbf{MCC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Root-mean-square_deviation}{\textbf{RMSE}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Sensitivity}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Specificity}}~$\uparrow$ \\
		\midrule
		$g^\text{GBM}$		& 0.775	& 0.857	& & 0.474 &	& 0.394 & & \\
		$g^\text{MGBM}$		& 0.763	& 0.846	& & 0.498 &	& 0.405 & & \\
		$g^\text{ANN}$		& 0.757	& 0.850	& 0.779 & 0.480 &	0.525 & 0.398 & 0.858 & \\
		$g^\text{XNN}$		& 0.758	& 0.851	& 0.781 & 0.479 &	0.528 & 0.397 & 0.867 & \\
		\bottomrule
		\label{tab:sim_acc}
	\end{tabular}
\end{table}

@@

For $g^\text{XNN}$, inherent interpretability manifests as plots of sparse $\gamma_k$ output layer weights, $n_k$ subnetwork ridge functions, and sparse $\beta_j$ weights in the bottom projection layer. Figure \ref{fig:sim_xnn_glob_ridge} provides detailed insights into $g^\text{XNN}$. \ref{fig:sim_xnn_glob_ridge} \textbf{A} displays the sparse $\gamma_k$ weights of the output layer, where only $n_k$ subnetworks with $k \in \{1, 4, 7, 8, 9\}$ are associated with large magnitude weights. The $n_k$ subnetwork ridge functions appear in \ref{fig:sim_xnn_glob_ridge} \textbf{B} as simplistic but distinctive functional forms. Color-coding between \ref{fig:sim_xnn_glob_ridge} \textbf{A} and \ref{fig:sim_xnn_glob_ridge} \textbf{B} visually reinforces the direct feed-forward relationship between the $n_k$ subnetworks and the $\gamma_k$ weights of the output layer.

\begin{figure}[H]
	\hskip-20pt\includegraphics[width=18cm]{img/sim_xnn_glob_ridge.png}
	\caption{\textbf{A.} Output layer $\gamma_k$ weights, \textbf{B.} corresponding $n_k$ ridge functions, and \textbf{C.} associated projection layer $\beta_j$ weights for $g^\text{XNN}$ on the simulated test data.}
	\label{fig:sim_xnn_glob_ridge}
\end{figure}

\noindent $n_k$ subnetworks are plotted across the output values of their associated $\sum_j\beta_{k,j}x_j$ projection layer hidden units, and color-coding between \ref{fig:sim_xnn_glob_ridge} \textbf{B} and \ref{fig:sim_xnn_glob_ridge} \textbf{C} link the $\beta_j$ weights to their $n_k$ subnetworks. Most of the heavily utilized $n_k$ subnetworks have sparse weights in their $\sum_j\beta_{k,j}x_j$ projection layer hidden units. In particular, subnetwork $n_1$ appears to be almost solely a function of $X_{\text{Friedman}, 3}$ and appears to exhibit the expected quadratic behavior for $X_{\text{Friedman}, 3}$.  Subnetworks $n_7$, $n_8$, and $n_9$ appear to be most associated with the globally important $X_{\text{Friedman}, 1}$ and $X_{\text{Friedman}, 2}$ features, likely betraying the effort required for $g^{\text{XNN}}$ to model the nonlinear $\text{sin}()$ function of the $X_{\text{Friedman}, 1}$ and $X_{\text{Friedman}, 2}$ product, and these subnetworks, especially $n_7$ and $n_8$, appear to display some noticeable sinusoidal characteristics. Subnetwork $n_4$ seems to be a linear combination of all the original input $X_j$ features, but does weigh the linear $X_{\text{Friedman},4}$ and $X_{\text{Friedman},5}$ terms roughly in the correct two-to-one ratio. As a whole, Figure \ref{fig:sim_xnn_glob_ridge} \textbf{A}, \textbf{B}, and \textbf{C} exhibit evidence that $g^\text{XNN}(\mathbf{X})$ has learned about the signal-generating function in Equation \ref{eq:sim} and the displayed information should help practitioners understand which original input $X_j$ features are weighed heavily in each $n_k$ subnetwork, and which $n_k$ subnetworks have a strong influence on $g^\text{XNN}(\mathbf{X})$ output.

@@

Table \ref{tab:mort_acc} shows that $g^\text{ANN}$ and $g^\text{XNN}$ noticeably outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data. This is at least partially due to the preprocessing required to present directly comparable post-hoc explainability results and to use neural networks and tensorflow, e.g., numerical encoding of categorical features and missing values. This preprocessing appears to hamstring some of the tree-based models' inherent capabilities. $g^\text{GBM}$ models trained on non-encoded data with missing values repeatedly produced area under the curve (AUC) values of $\sim0.81$ (not shown, but available in resources discussed in Subsection \ref{ssec:soft}).

\begin{table}[htb]
	\caption{Fit measures for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the mortgage test data. Arrows indicate the direction of improvement for each measure.}
\centering
\begin{tabular}{ccccccccc}
	\toprule
	\textbf{Model} & 
	\href{https://scikit-learn.org/stable/modules/model_evaluation.html#accuracy-score}{\textbf{Accuracy}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve}{\textbf{AUC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/F1_score}{\textbf{F1}}~$\uparrow$ & 
	\href{https://scikit-learn.org/stable/modules/model_evaluation.html#log-loss}{\textbf{Logloss}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Matthews_correlation_coefficient}{\textbf{MCC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Root-mean-square_deviation}{\textbf{RMSE}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Sensitivity}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Specificity}}~$\uparrow$ \\
	\midrule	
$g^\text{GBM}$		& 0.795	& 0.828	& & 0.252 &	& 0.276 & & \\
$g^\text{MGBM}$		& 0.765	& 0.814	& & 0.259 &	& 0.278 & & \\
$g^\text{ANN}$		& 0.865	& 0.871	& 0.474 & 0.231 &	0.418 & 0.262 & 0.624 & \\
$g^\text{XNN}$		& 0.869	& 0.868	& 0.468 & 0.233 &	0.409 & 0.263 & 0.593 & \\
\bottomrule
\label{tab:mort_acc}
\end{tabular}
\end{table}

\noindent Regardless of the fit differences between the two families of hypothesis models, the difference between the fit of constrained and unconstrained variants within the two types of models is small for the GBMs and negligible for ANNs, 3\% and < 1\% worse fit respectively, averaged across the metrics reported in Table \ref{tab:mort_acc}.

@@

For $g^\text{MGBM}$, intrinsic interpretability is evaluated with PD and ICE plots of mostly monotonic prediction behavior for several important $X_j$, and post-hoc Shapley explanation analysis is used to create global and local feature importance. Global Shapley feature importance for $g^\text{MGBM}$ on the mortgage test data is reported in Figure \ref{fig:mort_mgbm_glob}. $g^\text{MGBM}$ places high importance on LTV ratio, perhaps too high, and also weighs DTI ratio, property value, loan amount, and introductory rate period heavily in many of its predictions. Tree SHAP values are reported in the margin space, prior to the application of the logit link function, and the reported numeric values can be interpreted as the mean absolute impact of each $X_j$ on $g^\text{MGBM}(\mathbf{X})$ in the mortgage test data in the $g^\text{MGBM}$ margin space.

@@

\noindent The potential over-emphasis of LTV ratio, and the de-emphasis of income, likely an important feature from a business perspective, and the encoded no introductory rate period flag feature may also contribute to the decreased performance of $g^\text{MGBM}$ as compared to $g^\text{XNN}$.

Domain knowledge was used to positively constrain DTI ratio and LTV ratio and to negatively constrain income and the loan term flag under $g^\text{MGBM}$. The monotonicity constraints for DTI ratio and LTV ratio are confirmed for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data in Figure \ref{fig:mort_mgbm_glob_pdp_ice}. Both DTI ratio and LTV ratio display positive monotonic behavior at all selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for ICE and on average with PD. Because PD curves generally follow the patterns of the ICE curves for both features, it's also likely that no strong interactions are at play for DTI ratio and LTV ratio under $g^\text{MGBM}$. Of course, the monotonicity constraints themselves can dampen the effects of non-monotonic interactions under $g^\text{MGBM}$, even if they do exist in the data, and perhaps due to non-synthetic noise in the mortgage data, this rigidity could also play a role in the performance differences between $g^\text{MGBM}$ and $g^\text{XNN}$ in the mortgage data not observed in Table \ref{tab:sim_acc} for the simulated data.

PD and ICE are displayed with a histogram to highlight any sparse regions in an input feature's domain. Because most ML models will always issue a prediction on any datum with a correct schema, it's crucial to consider whether a given model learned enough about an observation to make an accurate prediction. Viewing PD and ICE along with a histogram is a convenient method to visually assess whether a prediction is reasonable and based on sufficient training data. DTI ratio and LTV ratio do appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as $g^\text{MGBM}$ appears to carry reasonable predictions learned from populous domains into the sparse domains of both features.

@@

Figure \ref{fig:mort_mgbm_glob_pdp_ice} also displays PD and ICE for the unconstrained feature property value. Unlike DTI ratio and LTV ratio, PD for property value does not always follow the patterns established by ICE curves. While PD shows monotonically increasing prediction behavior on average, apparently influenced by large predictions at extreme $g^{\text{MGBM}}(\mathbf{X})$ percentiles, ICE curves for individuals at the 40\textsuperscript{th} percentile of $g^{\text{MGBM}}(\mathbf{X})$, and lower, exhibit different prediction behavior with respect to property value. Some individuals at these lower percentiles display monotonically decreasing prediction behavior while others appear to show fluctuating prediction behavior. Property value is strongly right-skewed, with little data regarding high-value property from which $g^{\text{MGBM}}$ can learn. For the most part, reasonable predictions do appear to be carried from more densely populated regions to more sparsely populated regions. However, prediction fluctuations at lower $g^{\text{MGBM}}(\mathbf{X})$ percentiles are visible, and appear in a sparse region of property value. This divergence of PD and ICE can be indicative of an interaction affecting property value under $g^{\text{MGBM}}$ \cite{ice_plots}, and analysis by surrogate decision tree did show evidence of numerous potential interactions in lower predictions ranges of $g^{\text{MGBM}}(\mathbf{X})$ \cite{art_and_sci} (not shown, but available in resources discussed in Subsection \ref{ssec:soft}). However, fluctuations in ICE can also be caused by overfitting or leakage of strong non-monotonic signal from important constrained features into the modeled behavior of non-constrained features.

@@

The capability of $g^\text{XNN}$ to model nonlinear phenomenon and high-degree interactions, and to do so in an interpretable manner, is on display in Figure \ref{fig:mort_xnn_glob_ridge}. \ref{fig:mort_xnn_glob_ridge} \textbf{A} presents the sparse $\gamma_k$ weights of the $g^\text{XNN}$ output layer in which the $n_k$ subnetworks with $k \in \{0,1,2,3,5,8,9\}$ have large magnitude weights and $n_k$ subnetworks, $k \in \{4,6,7\}$, have small or near-zero magnitude weights. Distinctive ridge functions that feed into those large magnitude $\gamma_k$ weights are highlighted in \ref{fig:mort_xnn_glob_ridge} \textbf{B} and color-coded to pair with their corresponding $\gamma_k$ weight. As in the Subsection \ref{ssec:xnn_sim}, $n_k$ ridge function plots vary with the output of the corresponding projection layer $\sum_j\beta_{k,j}x_j$ hidden unit, with weights displayed in matching colors in \ref{fig:mort_xnn_glob_ridge} \textbf{C}. In both the simulated and mortgage data, $n_k$ ridge functions appear to be elementary functional forms that the output layer learns to combine to generate accurate predictions, reminiscent of basis functions for the modeled space. \ref{fig:mort_xnn_glob_ridge} \textbf{C} displays the sparse $\beta_j$ weights of the projection layer $\sum_j\beta_{k,j}x_j$ hidden units that are associated with each $n_k$ subnetwork ridge function. For instance, subnetwork $n_3$ is influenced by large weights for LTV ratio, no introductory rate period flag, and introductory rate period, whereas subnetwork $n_9$ is nearly completely dominated by the weight for income.

\begin{figure}[htb]
\hskip-20pt\includegraphics[width=18cm]{img/mort_xnn_glob_ridge.png}
\caption{\textbf{A.} Output layer $\gamma_k$ weights, \textbf{B.} corresponding $n_k$ ridge functions, and \textbf{C.} associated projection layer $\beta_j$ weights for $g^\text{XNN}$ on the mortgage test data.}
\label{fig:mort_xnn_glob_ridge}
\end{figure}

>>>>

% TODO:
% EDITING:
% - Structure ------------------------- x 
% - External links
% - Internal links
%   - Sections/subsections/appendices - x
%   - References ---------------------- x
%   - Tables/figures
%   - footnotes ----------------------- x
% - Abbreviations --------------------- x
% - Bibliography 
%   - Nick/Bryce citation
% - Notation
% - Tables/figures
% - Grammar and spelling
%   - Hyphenation
%   - "metric" vs. "measure"
% CONTENT:
% - sim data: add error metrics, confirm statements about results ------------------------------------------------------------------------------------------- x
% - sim data xnn: update figure: connections, feed-forward direction, make subfigure reference match discrimination measure table --- x
% - mort data: add error metrics, confirm statements about results ----------------------------------------------------------------------------------------- x
% - mort data: update figure: connections, feed-forward direction, make subfigure reference match discrimination measure table ------- x
% - make subfigure reference match XNN charts ---------------------------------------------------------------------------------------------------------------- x

@@

Fit comparisons between unconstrained and constrained models and XNN interpretabililty results are discussed in Subsections \ref{ssec:c_v_uc_sim} and \ref{ssec:xnn_sim}. As model training and assessment on the simulated data is a rough validation exercise meant to showcase expected results on data with known characteristics, and given that most of the techniques in the proposed workflow are already used widely or have been validated elsewhere, reporting of simulated data results in the main text will focus mostly on fit measures for the considered data and model architectures and on the more novel $g^\text{XNN}$ interpretability results. The bulk of the post-hoc explanation and discrimination testing results for the simulated data are left to Appendix \ref{a_sec:res_sim}.

@@

Table \ref{tab:sim_acc} presents a variety of fit measures for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data. $g^\text{XNN}$ exhibits the best performance, but the models exhibit only a fairly small range of fit results. Interpretability and explainability benefits of the constrained models appear to come at little cost to overall model performance, or in the case of $g^\text{ANN}$ and $g^\text{XNN}$, no cost at all. For the displayed measures, $g^\text{MGBM}$ performs $\sim$2\% worse then $g^\text{GBM}$. $g^\text{XNN}$ performs $\sim$0.5\% better on average than $g^\text{XNN}$, and $g^\text{XNN}$ actually shows slightly better fit than $g^\text{ANN}$ across all fit measures except specificity. Fit measures that require a probability cutoff are taken at the best F1 threshold for each model.

\begin{table}[H]
	\caption{Fit measures for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the simulated test data. Arrows indicate the direction of improvement for each measure.}
	\centering
	\begin{tabular}{ccccccccc}
		\toprule
		\textbf{Model} & 
		\href{https://scikit-learn.org/stable/modules/model_evaluation.html#accuracy-score}{\textbf{Accuracy}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve}{\textbf{AUC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/F1_score}{\textbf{F1}}~$\uparrow$ & 
		\href{https://scikit-learn.org/stable/modules/model_evaluation.html#log-loss}{\textbf{Logloss}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Matthews_correlation_coefficient}{\textbf{MCC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Root-mean-square_deviation}{\textbf{RMSE}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Sensitivity}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Specificity}}~$\uparrow$ \\
		\midrule
		$g^\text{GBM}$		& 0.757 	& 0.847 	& 0.779 & 0.486 & 0.525 & 0.400 & 0.858 & 0.657 \\
		$g^\text{MGBM}$	& 0.744 	& 0.842 	& 0.771 & 0.502 & 0.504 & 0.407 & 0.864 & 0.625 \\
		$g^\text{ANN}$		& 0.757	& 0.850	& 0.779 & 0.480 & 0.525 & 0.398 & 0.858 & 0.657 \\
		$g^\text{XNN}$		& 0.758	& 0.851	& 0.781 & 0.479 & 0.528 & 0.397 & 0.867 & 0.648 \\
		\bottomrule
		\label{tab:sim_acc}
	\end{tabular}
\end{table}

@@

For $g^\text{XNN}$, inherent interpretability manifests as plots of sparse $\gamma_k$ output layer weights, $n_k$ subnetwork ridge functions, and sparse $\beta_j$ weights in the bottom projection layer. Figure \ref{fig:sim_xnn_glob_ridge} provides detailed insights into $g^\text{XNN}$. \ref{fig:sim_xnn_glob_ridge}\textbf{a} displays the sparse $\gamma_k$ weights of the output layer, where only $n_k$ subnetworks with $k \in \{1, 4, 7, 8, 9\}$ are associated with large magnitude weights. The $n_k$ subnetwork ridge functions appear in \ref{fig:sim_xnn_glob_ridge}\textbf{b} as simplistic but distinctive functional forms. Color-coding between \ref{fig:sim_xnn_glob_ridge}\textbf{a} and \ref{fig:sim_xnn_glob_ridge}\textbf{b} visually reinforces the direct feed-forward relationship between the $n_k$ subnetworks and the $\gamma_k$ weights of the output layer.

\begin{figure}[H]
	\hskip-20pt\includegraphics[width=18cm]{img/sim_xnn_glob_ridge.png}
	\caption{Output layer $\gamma_k$ weights, corresponding $n_k$ ridge functions, and associated projection layer $\beta_j$ weights for $g^\text{XNN}$ on the simulated test data.}
	\label{fig:sim_xnn_glob_ridge}
\end{figure}

\noindent $n_k$ subnetworks are plotted across the output values of their associated $\sum_j\beta_{k,j}x_j$ projection layer hidden units, and color-coding between \ref{fig:sim_xnn_glob_ridge}\textbf{b} and \ref{fig:sim_xnn_glob_ridge}\textbf{c} link the $\beta_j$ weights to their $n_k$ subnetworks. Most of the heavily utilized $n_k$ subnetworks have sparse weights in their $\sum_j\beta_{k,j}x_j$ projection layer hidden units. In particular, subnetwork $n_1$ appears to be almost solely a function of $X_{\text{Friedman}, 3}$ and appears to exhibit the expected quadratic behavior for $X_{\text{Friedman}, 3}$.  Subnetworks $n_7$, $n_8$, and $n_9$ appear to be most associated with the globally important $X_{\text{Friedman}, 1}$ and $X_{\text{Friedman}, 2}$ features, likely betraying the effort required for $g^{\text{XNN}}$ to model the nonlinear $\text{sin}()$ function of the $X_{\text{Friedman}, 1}$ and $X_{\text{Friedman}, 2}$ product, and these subnetworks, especially $n_7$ and $n_8$, appear to display some noticeable sinusoidal characteristics. Subnetwork $n_4$ seems to be a linear combination of all the original input $X_j$ features, but does weigh the linear $X_{\text{Friedman},4}$ and $X_{\text{Friedman},5}$ terms roughly in the correct two-to-one ratio. As a whole, Figure \ref{fig:sim_xnn_glob_ridge}\textbf{a}, \textbf{b}, and \textbf{c} exhibit evidence that $g^\text{XNN}(\mathbf{X})$ has learned about the signal-generating function in Equation \ref{eq:sim} and the displayed information should help practitioners understand which original input $X_j$ features are weighed heavily in each $n_k$ subnetwork, and which $n_k$ subnetworks have a strong influence on $g^\text{XNN}(\mathbf{X})$ output. See Appendix \ref{a_ssec:xnn} for details regarding general XNN architecture.

@@

Table \ref{tab:mort_acc} shows that $g^\text{ANN}$ and $g^\text{XNN}$ noticeably outperform $g^\text{GBM}$ and $g^\text{MGBM}$ on the mortgage data for most of the fit measures. This is at least partially due to the preprocessing required to present directly comparable post-hoc explainability results and to use neural networks and tensorflow, e.g., numerical encoding of categorical features and missing values. This preprocessing appears to hamstring some of the tree-based models' inherent capabilities. $g^\text{GBM}$ models trained on non-encoded data with missing values repeatedly produced area under the curve (AUC) values of $\sim0.81$ (not shown, but available in resources discussed in Subsection \ref{ssec:soft}).

\begin{table}[htb]
	\caption{Fit measures for $g^\text{GBM}$, $g^\text{MGBM}$, $g^\text{ANN}$, and $g^\text{XNN}$ on the mortgage test data. Arrows indicate the direction of improvement for each measure.}
\centering
\begin{tabular}{ccccccccc}
	\toprule
	\textbf{Model} & 
	\href{https://scikit-learn.org/stable/modules/model_evaluation.html#accuracy-score}{\textbf{Accuracy}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Receiver_operating_characteristic#Area_under_the_curve}{\textbf{AUC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/F1_score}{\textbf{F1}}~$\uparrow$ & 
	\href{https://scikit-learn.org/stable/modules/model_evaluation.html#log-loss}{\textbf{Logloss}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Matthews_correlation_coefficient}{\textbf{MCC}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Root-mean-square_deviation}{\textbf{RMSE}}~$\downarrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Sensitivity}}~$\uparrow$ & \href{https://en.wikipedia.org/wiki/Sensitivity_and_specificity}{\textbf{Specificity}}~$\uparrow$ \\
	\midrule	
$g^\text{GBM}$		& 0.795	& 0.828	& 0.376 & 0.252 & 0.314 & 0.276 & 0.634 & 0.813 \\
$g^\text{MGBM}$	& 0.765	& 0.814	& 0.362 & 0.259 & 0.305 & 0.278 & 0.684 & 0.773 \\
$g^\text{ANN}$		& 0.865	& 0.871	& 0.474 & 0.231 & 0.418 & 0.262 & 0.624 & 0.891 \\
$g^\text{XNN}$		& 0.869	& 0.868	& 0.468 & 0.233 & 0.409 & 0.263 & 0.594 & 0.898 \\
\bottomrule
\label{tab:mort_acc}
\end{tabular}
\end{table}

\noindent Regardless of the fit differences between the two families of models, the difference between the constrained and unconstrained variants within the two types of models is small for the GBMs and smaller for the ANNs, $\sim$3.5\% and $\sim$1\% worse fit respectively, averaged across the measures in Table \ref{tab:mort_acc}.

@@

For $g^\text{MGBM}$, intrinsic interpretability is evaluated with PD and ICE plots of mostly monotonic prediction behavior for several important $X_j$, and post-hoc Shapley explanation analysis is used to create global and local feature importance. Global Shapley feature importance for $g^\text{MGBM}$ on the mortgage test data is reported in Figure \ref{fig:mort_mgbm_glob}. $g^\text{MGBM}$ places high importance on LTV ratio, perhaps too high, and also weighs DTI ratio, property value, loan amount, and introductory rate period heavily in many of its predictions. Tree SHAP values are reported in the margin space, prior to the application of the logit link function, and the reported numeric values can be interpreted as the mean absolute impact of each $X_j$ on $g^\text{MGBM}(\mathbf{X})$ in the mortgage test data in the $g^\text{MGBM}$ margin space. The potential over-emphasis of LTV ratio, and the de-emphasis of income, likely an important feature from a business perspective, and the encoded no introductory rate period flag feature may also contribute to the decreased performance of $g^\text{MGBM}$ as compared to $g^\text{XNN}$.

@@

Domain knowledge was used to positively constrain DTI ratio and LTV ratio and to negatively constrain income and the loan term flag under $g^\text{MGBM}$. The monotonicity constraints for DTI ratio and LTV ratio are confirmed for $g^\text{MGBM}(\mathbf{X})$ on the mortgage test data in Figure \ref{fig:mort_mgbm_glob_pdp_ice}. Both DTI ratio and LTV ratio display positive monotonic behavior at all selected percentiles of $g^\text{MGBM}(\mathbf{X})$ for ICE and on average with PD. Because PD curves generally follow the patterns of the ICE curves for both features, it's also likely that no strong interactions are at play for DTI ratio and LTV ratio under $g^\text{MGBM}$. Of course, the monotonicity constraints themselves can dampen the effects of non-monotonic interactions under $g^\text{MGBM}$, even if they do exist in the data. Perhaps due to the presence of important non-monotonic interactions (e.g., LTV ratio and the no introductory rate period flag, see Figure \ref{fig:mort_xnn_glob_ridge}). This rigidity could also play a role in the performance differences between $g^\text{MGBM}$ and $g^\text{XNN}$ in the mortgage data not observed for the simulated data, wherein strong interactions appear to be between features with the same monotonicity constraints (e.g., $X_{\text{Friedman},1}$ and $X_{\text{Friedman},2}$, see Figure \ref{fig:sim_xnn_glob_ridge}).

PD and ICE are displayed with a histogram to highlight any sparse regions in an input feature's domain. Because most ML models will always issue a prediction on any instance with a correct schema, it's crucial to consider whether a given model learned enough about an observation to make an accurate prediction. Viewing PD and ICE along with a histogram is a convenient method to visually assess whether a prediction is reasonable and based on sufficient training data. DTI ratio and LTV ratio do appear to have sparse regions in their univariate distributions. The monotonicity constraints likely play to the advantage of $g^\text{MGBM}$ in this regard, as $g^\text{MGBM}$ appears to carry reasonable predictions learned from populous domains into the sparse domains of both features.

@@

Figure \ref{fig:mort_mgbm_glob_pdp_ice} also displays PD and ICE for the unconstrained feature property value. Unlike DTI ratio and LTV ratio, PD for property value does not always follow the patterns established by ICE curves. While PD shows monotonically increasing prediction behavior on average, apparently influenced by large predictions at extreme $g^{\text{MGBM}}(\mathbf{X})$ percentiles, ICE curves for individuals at the 40\textsuperscript{th} percentile of $g^{\text{MGBM}}(\mathbf{X})$, and lower, exhibit different prediction behavior with respect to property value. Some individuals at these lower percentiles display monotonically decreasing prediction behavior while others appear to show fluctuating prediction behavior. Property value is strongly right-skewed, with little data regarding high-value property from which $g^{\text{MGBM}}$ can learn. For the most part, reasonable predictions do appear to be carried from more densely populated regions to more sparsely populated regions. However, prediction fluctuations at lower $g^{\text{MGBM}}(\mathbf{X})$ percentiles are visible, and appear in a sparse region of property value. This divergence of PD and ICE can be indicative of an interaction affecting property value under $g^{\text{MGBM}}$ \cite{ice_plots}, and analysis by surrogate decision tree did show evidence of numerous potential interactions in lower predictions ranges of $g^{\text{MGBM}}(\mathbf{X})$ \cite{art_and_sci} (not shown, but available in resources discussed in Subsection \ref{ssec:soft}). However, fluctuations in ICE can also be caused by overfitting or by leakage of strong non-monotonic signal from important constrained features into the modeled behavior of non-constrained features.

@@

The capability of $g^\text{XNN}$ to model nonlinear phenomenon and high-degree interactions, and to do so in an interpretable manner, is on display in Figure \ref{fig:mort_xnn_glob_ridge}. \ref{fig:mort_xnn_glob_ridge}\textbf{a} presents the sparse $\gamma_k$ weights of the $g^\text{XNN}$ output layer in which the $n_k$ subnetworks with $k \in \{0,1,2,3,5,8,9\}$ have large magnitude weights and $n_k$ subnetworks, $k \in \{4,6,7\}$, have small or near-zero magnitude weights. Distinctive ridge functions that feed into those large magnitude $\gamma_k$ weights are highlighted in \ref{fig:mort_xnn_glob_ridge}\textbf{b} and color-coded to pair with their corresponding $\gamma_k$ weight. As in the Subsection \ref{ssec:xnn_sim}, $n_k$ ridge function plots vary with the output of the corresponding projection layer $\sum_j\beta_{k,j}x_j$ hidden unit, with weights displayed in matching colors in \ref{fig:mort_xnn_glob_ridge}\textbf{c}. In both the simulated and mortgage data, $n_k$ ridge functions appear to be elementary functional forms that the output layer learns to combine to generate accurate predictions, reminiscent of basis functions for the modeled space. \ref{fig:mort_xnn_glob_ridge}\textbf{c} displays the sparse $\beta_j$ weights of the projection layer $\sum_j\beta_{k,j}x_j$ hidden units that are associated with each $n_k$ subnetwork ridge function. For instance, subnetwork $n_3$ is influenced by large weights for LTV ratio, no introductory rate period flag, and introductory rate period, whereas subnetwork $n_9$ is nearly completely dominated by the weight for income. See Appendix \ref{a_ssec:xnn} for details regarding general XNN architecture.

\begin{figure}[htb]
\hskip-20pt\includegraphics[width=18cm]{img/mort_xnn_glob_ridge.png}
\caption{Output layer $\gamma_k$ weights, corresponding $n_k$ ridge functions, and associated projection layer $\beta_j$ weights for $g^\text{XNN}$ on the mortgage test data.}
\label{fig:mort_xnn_glob_ridge}
\end{figure}
